---
title: "first_draft"
author: "I. Tulenheimo, L. Toivanen"
format: pdf
editor: visual
---

# TODO:

-   Selvitä smoking datan "highest_qualification" sarakkeen tutkintojen "paremmuus järjestys", eli mikä tutkinto huonoin ja mikä paras (jotta voidaan esim. numeroida tutkinnot huonoimmasta parhaimpaan).

Let us first install missing libraries (if any) and load the library containing the smoking data.

```{r}
# Let us define a list of needed packages, compare it to the output from installed.packages()[,"Package"] and install the missing packages:
list.of.packages <- c("openintro", "ggplot2")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load the library containing the smoking data and other libraries needed
library(ggplot2)
library(openintro)
smoking_df = smoking
```

## Visualise the smoking data

Let's first have a quick look at the data:

```{r}
ggplot(smoking, aes(x = amt_weekends)) +
  geom_histogram(binwidth = 5)

ggplot(smoking, aes(x = amt_weekdays)) +
  geom_histogram(binwidth = 5)

ggplot(smoking, aes(x = gender, fill = smoke)) +
  geom_bar(position = "fill")

ggplot(smoking, aes(x = marital_status, fill = smoke)) +
  geom_bar(position = "fill")
```

Let's look at the distribution of smokers w.r.t. highest qualification:

```{r}
ggplot(smoking, aes(x = highest_qualification, fill = smoke)) +
  geom_bar(position = "fill") + 
  # Rotate x-axis labels
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

Let us also check the number of data points for each qualification:

```{r}
# Use table function to count the number of rows for each degree
degree_counts <- table(smoking_df$highest_qualification)

# Print the result
print(degree_counts)
```

### Potential research questions

Looking at the histogram plot above, couple of questions arise:

-   What is the trend of smoking as measured by hihgest qualification (i.e does the number of smokers increase or decrease with respect to the education level)?

-   Given the number of smokers for the first seven qualifications (when sorted from lowest to highest degree), how many smokers are there with the highest degree of qualification/education?

### Discussion of potential models

To answer our two questions of interest, we could try to fit for example the following models:

-   (simple) non hierarchical linear model

-   and hierarchical linear model.

## Data preprocessing

Before implementing the actual models, we must do some preprocessing of the data.

First, let us drop the rows that have no information about the education, smoking status or gender, since those are our main interests.

```{r}
# Count the number of rows before dropping
rows_before <- nrow(smoking)

# Drop rows with "NA" in the "highest_qualification" or "smoke" columns
smoking_df <- smoking_df[complete.cases(smoking_df$highest_qualification, smoking_df$smoke, smoking_df$gender), , drop = FALSE]

# Count the number of rows after dropping
rows_after <- nrow(smoking_df)

# Calculate the number of rows dropped
rows_dropped <- rows_before - rows_after

# Print the result
print(smoking_df)
cat("Number of before filtering:", rows_before, "\n")
cat("Number of after filtering:", rows_after, "\n")
cat("Number of rows dropped:", rows_dropped, "\n")
```

Second, we will need to map the string representation of degrees to numeric values. We will map the degrees from lowest to highest (from 1 to 8):

```{r}
# Define mapping of degrees to numeric values
degree_mapping <- c("No Qualification" = 1, "Other/Sub Degree" = 2, "GCSE/CSE" = 3, "GCSE/O Level" = 4, "ONC/BTEC" = 5, "A Levels" = 6, "Higher/Sub Degree" = 7, "Degree" = 8)

# Use factor to map the degrees to numeric values
smoking_df$highest_qualification_num <- as.numeric(factor(smoking_df$highest_qualification, levels = names(degree_mapping)))

# Print the result
print(smoking_df)
```

Degree order from highest to lowest:

1.  No Qualification

2.  Other/Sub Degree

3.  GCSE/CSE

4.  GCSE/O Level

5.  ONC/BTEC

6.  A Levels

7.  Higher/Sub Degree

8.  Degree

```{r}
ggplot(smoking_df, aes(x = highest_qualification_num, fill = smoke)) +
  geom_bar(position='fill')

ggplot(smoking, aes(x = age, fill = smoke)) +
  geom_histogram(position='fill', bnis=30)

ggplot(smoking_df, aes(x = highest_qualification_num, fill = smoke)) +
  geom_bar()

ggplot(smoking, aes(x = age, fill = smoke)) +
  geom_histogram(bnis=30)
```

Third, let us drop the columns that are not needed. Let us have a simple start and only preserve the information about the gender, education, and smoking status.

```{r}
# List of columns to preserve
columns_to_preserve <- c("gender", "highest_qualification", "highest_qualification_num", "smoke")

# Select the columns
smoking_df <- smoking_df[, columns_to_preserve, drop = FALSE]

smoking_df
```

Playground:

```{r}
# Use table function to count the number of rows for each degree
degree_counts <- table(smoking_df$highest_qualification_num)

# Print the result
print(degree_counts)
```

The `echo: false` option disables the printing of code (only output is displayed).
